Trends from 30-year observations of downward solar irradiance in Thessaloniki, Greece

### 2023-12-18 MDPI_submission.Rmd
Commit 2023-12-18 20:14

## some vars to use later
alltrend   <- signif(LT[DATA == "All sky cond.",    slope], 2)
cleartrend <- signif(LT[DATA == "Clear sky cond.",  slope], 2)
cloudtrend <- signif(LT[DATA == "Cloudy sky cond.", slope], 2)
LTminSSig  <- signif(min(LT$slope.stat_sig), 4)

@@

We calculated the linear trends of SDR, from the departures of the mean daily values from the daily climatology and for the three sky conditions (Table\ \ref{tab:trendtable}).
In Figure\ \ref{fig:trendALL} we present only the time series under all-sky conditions; the plots for clear-sky and cloud-sky conditions, are shown in the Appendix (Figures\ \ref{fig:trendCLEAR} and\  \ref{fig:trendCLOUD}).
In the studied period, there is no
significant break or change in the variability pattern of the time
series.
We observe a positive trend for all-sky conditions
($`r alltrend`\,\%/y$), a very close but smaller trend for clear-skies ($`r cleartrend`\,\%/y$) and a negative weaker trend for cloudy-skies ($`r cloudtrend`\,\%/y$).
In the studied period, there is no significant break or change in the variability of the time series.
Other studies for the European region reported a change of the SDR slope, around 1980 [@Wild2021; @Yuan2021; @Ohmura2009], a few years before the start of our records.
It is interesting to note, that for the observations period, the trend of the TSI is
$`r format(TSIslopeAll[1,1], scientific = F, digits = 2)`\,\%/y$,
and thus we can not attribute any major effect on SDR trend to Solar variability.

<!-- \footnotesize -->
<!-- ```{r trendtable2, echo=F} -->
<!-- panderOptions("table.alignment.default", "center") -->
<!-- pander(LT1, -->
<!--        caption = paste0("\\label{tab:trendtable2}Trends in SDR daily mean for different sky conditions for the period", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), ". For all cases the statistical significance is above $", LTminSSig, "\\,\\%$.")) -->
<!-- ``` -->
<!-- \normalsize -->

```{r trendtable, echo=F}

# names(LT1)
# "Sky conditions", "Trend [%/year]", "Trend S.E.", "Pearson cor.", "Trend [W/year]", "Days" 
x_latex <- knitr::kable(LT1,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = FALSE,
                        col.names   = c("Sky conditions", "Trend [\\%/year]", "Trend S.E. ($2\\sigma$)", "Pearson cor.", "Trend [W/year]", "Days"),
                        format.args = list(digits = 3),
                        linesep     = "",
                        align       = "crrrrr",
                        centering   = FALSE,
                        table.envir = "table",
                        position    = "H",
                        caption     = paste0("Trends in SDR daily means for different sky conditions for the period ", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), "."))

@@

For cloudy-sky conditions (Figure\ \ref{fig:szatrends-3}), we can not discern any significant dependence of the SDR trend with SZA as the variability of irradiance is
dominated by the cloud effects.
For SZAs $25^\circ$ -- $75^\circ$, the trends range within about $\pm 0.2\,\%/y$,
the trends are variable and for most SZAs insignificant ranging between $0$ and $0.2\,\%/y$.
The smooth increase of the trend at SZAs larger than $75^\circ$ is probably associated with stronger attenuation by
clouds under oblique incidence angles which results in smaller
variability.

@@

## numeric precision
spp$slope          <- signif(spp$slope,          3)
spp$slope.p        <- signif(spp$slope.p,        3)
spp$slope.stat_sig <- signif(spp$slope.stat_sig, 3)
spp$Rsqrd          <- signif(spp$Rsqrd,          3)
spp$RsqrdAdj       <- NULL
spp$Rsqrd          <- NULL
spp$cor.t          <- NULL
spp$slope.stat_sig <- NULL
spp$N_eff          <- NULL

@@

spp <- rename(spp, "Trend [%/year]" = slope)
spp <- rename(spp, "Sky condition"  = DATA )

# names(spp) <- c("Trend [%/year]", "Trend p-value", "Sky condition", "Season", "Statistical signif. [%]")

@@

Similarly to the long term trends from daily means of SDR discussed above, we have calculated the trend of the anomalies for the three different sky conditions, and for each season of the year, using the corresponding mean monthly values
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
For all-sky conditions the trend in SDR in winter is the largest
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$),
followed by the trend in autumn
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$,
a value close to the long term trend) both statistically significant above the $99\,\%$ confidence level.
In spring and summer, the trends are much smaller and of lesser statistical significance.
These seasonal differences indicate a possible relation of the trends in SDR to trends of clouds during winter and autumn.
For clear-skies, the trend in winter is $`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$, larger than for all-skies ($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$), which is another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost complementary to each other, particularly for winter and autumn, where the signal is stronger.
During spring and summer the statistical significance is very low and the actual trend too small for a meaningful comparison.

@@

## arrange columns
# spp <- spp[,c(
#       "Sky condition",
#       "Season",
#       "Trend [%/year]",
#       # "Trend p-value",
#       "Statistical signif. [%]"
#       )]

@@

## use kable
x_latex <- knitr::kable(spp2,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = TRUE,
                        # format.args = list(digits = 3),
                        linesep     = "",
                        align       = "ccrr",
                        centering   = FALSE,
                        table.envir = "table",
                        # position    = "H",
                        caption  = "SDR linear trends of monthly anomalies for each season of the year.")

>>>>

## some vars to use later
alltrend   <- signif(LT[DATA == "All sky cond.",    slope], 2)
cleartrend <- signif(LT[DATA == "Clear sky cond.",  slope], 1)
cloudtrend <- signif(LT[DATA == "Cloudy sky cond.", slope], 2)
LTminSSig  <- signif(min(LT$slope.stat_sig), 4)

@@

We calculated the linear trends of SDR, from the departures of the mean daily values from the daily climatology and for the three sky conditions (Table\ \ref{tab:trendtable}).
In Figure\ \ref{fig:trendALL} we present only the time series under all-sky conditions; the plots for clear-sky and cloud-sky conditions, are shown in the Appendix (Figures\ \ref{fig:trendCLEAR} and\  \ref{fig:trendCLOUD}).
In the studied period, there is no
significant break or change in the variability pattern of the time
series.
The linear trends in all three datasets are positive and around
$0.4\,\%/y$ for all-sky and cloudy-sky conditions, while for clear-skies the
trend is much smaller (\~$`r cloudtrend`\,\%/y$). 
The linear trends were calculated
taking into account the autocorrelation of the time series and all three
are statistically significant at least at the $95\,\%$ confidence level, as
they are larger than the corresponding $2\sigma$ errors.
The clear-sky trend is
very small suggesting a small effect from aerosols and water vapor which
are the dominant factors of the SDR variability. 
In contrast the large
positive trend of SDR under cloudy skies can be attributed to reduction
in cloud cover and/or optical depth. 
Lack of continuous cloud
observations that could support these findings does not allow drawing
firm conclusions. However, there are indications that the total
cloud-cover as inferred from the ERA5 analysis for the grid point of
Thessaloniki is decreasing over the period of study.

<!-- ($`r alltrend`\,\%/y$), a very close but smaller trend for clear-skies ($`r cleartrend`\,\%/y$) and a negative weaker trend for cloudy-skies ($`r cloudtrend`\,\%/y$). -->
The all-sky trend is similar to the one reported in @Bais2013
from a ten-year shorter dataset suggesting that the tendency of SDR in
Thessaloniki is systematic.
Other studies for the European region reported a change in the SDR trend
around 1980 from negative to positive
with comparable magnitude [@Wild2021; @Yuan2021; @Ohmura2009], 
well before the start of our records.
However, the
trends reported here for the three datasets are in accordance with the
widely accepted solar radiation brightening over Europe.
For the period
of our observations the trend in the TSI is negligible
($`r format(TSIslopeAll[1,1], scientific = F, digits = 2)`\,\%/y$),
and thus we cannot attribute any significant effect
on the SDR trend to solar variability.

```{r trendtable, echo=F}
# names(LT1)
# "Sky conditions", "Trend [%/year]", "Trend S.E.", "Pearson cor.", "Trend [W/year]", "Days" 
x_latex <- knitr::kable(LT1,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = FALSE,
                        col.names   = c("Sky conditions", "Trend [\\%/year]", "Trend S.E. ($2\\sigma$)", "Pearson cor.", "Trend [W/year]", "Days"),
                        format.args = list(digits = 3),
                        linesep     = "",
                        align       = "crrrrr",
                        centering   = FALSE,
                        table.envir = "table",
                        position    = "H",
                        caption     = paste0("Trends in SDR daily means for different sky conditions for the period ", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), "."))

@@

For cloudy-sky conditions (Figure\ \ref{fig:szatrends-3}), we can not discern any significant dependence of the SDR trend with SZA as the variability of irradiance is
dominated by the cloud effects.
For SZAs $25^\circ$ -- $75^\circ$,
the trends are variable and for most SZAs insignificant ranging between $0$ and $0.2\,\%/y$.
The smooth increase of the trend at SZAs larger than $\sim 75^\circ$ is probably associated with stronger attenuation by
clouds under oblique incidence angles which results in smaller
variability.

@@

## numeric precision
spp$slope          <- signif(spp$slope,          3)
spp$slope.p        <- signif(spp$slope.p,        3)
spp$slope.stat_sig <- signif(spp$slope.stat_sig, 3)
spp$Rsqrd          <- signif(spp$Rsqrd,          3)
spp$cor.estimate   <- signif(spp$cor.estimate,   3)
spp$RsqrdAdj       <- NULL
spp$Rsqrd          <- NULL
spp$cor.t          <- NULL
spp$slope.stat_sig <- NULL
spp$N_eff          <- NULL
spp$conf_2.5       <- NULL
spp$conf_97.5      <- NULL
spp$t_eff          <- NULL
spp$t_eff_cri      <- NULL

@@

spp <- rename(spp, "Trend [%/year]" = slope       )
spp <- rename(spp, "Sky condition"  = DATA        )
spp <- rename(spp, "Pearson cor."   = cor.estimate)

## arrange columns
spp <- spp[,c(
      "Sky condition",
      "Season",
      "Trend [%/year]",
      "Pearson cor.",
      "Trend p-value"
      )]

@@

Similarly to the long term trends from daily means of SDR discussed above, we have calculated the trend for the three sky conditions, and for each season of the year, using the corresponding mean monthly anomalies
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
For all-sky conditions the trend in SDR in winter is the largest
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$),
followed by the trend in autumn
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$,
a value close to the long term trend) both statistically significant above the $99\,\%$ confidence level.
In spring and summer, the trends are much smaller and of lesser statistical significance.
These seasonal differences indicate a possible relation of the trends in SDR to trends of clouds during winter and autumn.
For clear-skies, the trend in winter is $`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$, larger than for all-skies ($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$), which is another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost complementary to each other, particularly for winter and autumn, where the signal is stronger.
During spring and summer the statistical significance is very low and the actual trend too small for a meaningful comparison.

@@

@@

## use kable
x_latex <- knitr::kable(spp2,
                        format      = "latex",
                        booktabs    = TRUE,
                        escape      = TRUE,
                        # format.args = list(digits = 3),
                        linesep     = "",
                        align       = "ccrrr",
                        centering   = FALSE,
                        table.envir = "table",
                        # position    = "H",
                        caption  = "SDR linear trends of monthly anomalies for each season of the year.")

### 2023-12-18 MDPI_submission.tex
Commit 2023-12-18 20:14

We calculated the linear trends of SDR, from the departures of the mean
daily values from the daily climatology and for the three sky conditions
(Table~\ref{tab:trendtable}). In Figure~\ref{fig:trendALL} we present
only the time series under all-sky conditions; the plots for clear-sky
and cloud-sky conditions, are shown in the Appendix
(Figures~\ref{fig:trendCLEAR} and~ \ref{fig:trendCLOUD}). In the studied
period, there is no significant break or change in the variability
pattern of the time series. We observe a positive trend for all-sky
conditions (\(0.38\,\%/y\)), a very close but smaller trend for
clear-skies (\(0.097\,\%/y\)) and a negative weaker trend for
cloudy-skies (\(0.41\,\%/y\)). In the studied period, there is no
significant break or change in the variability of the time series. Other
studies for the European region reported a change of the SDR slope,
around 1980 \citep{Wild2021, Yuan2021, Ohmura2009}, a few years before
the start of our records. It is interesting to note, that for the
observations period, the trend of the TSI is \(-0.00024\,\%/y\), and
thus we can not attribute any major effect on SDR trend to Solar
variability.

@@

For cloudy-sky conditions (Figure~\ref{fig:szatrends-3}), we can not
discern any significant dependence of the SDR trend with SZA as the
variability of irradiance is dominated by the cloud effects. For SZAs
\(25^\circ\) -- \(75^\circ\), the trends range within about
\(\pm 0.2\,\%/y\), the trends are variable and for most SZAs
insignificant ranging between \(0\) and \(0.2\,\%/y\). The smooth
increase of the trend at SZAs larger than \(75^\circ\) is probably
associated with stronger attenuation by clouds under oblique incidence
angles which results in smaller variability.

@@

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend of the anomalies for the three
different sky conditions, and for each season of the year, using the
corresponding mean monthly values (Figure~\ref{fig:seasonalALL} and
Table~\ref{tab:trendseasontable}). For all-sky conditions the trend in
SDR in winter is the largest (\(0.7\,\%/y\)), followed by the trend in
autumn (\(0.42\,\%/y\), a value close to the long term trend) both
statistically significant above the \(99\,\%\) confidence level. In
spring and summer, the trends are much smaller and of lesser statistical
significance. These seasonal differences indicate a possible relation of
the trends in SDR to trends of clouds during winter and autumn. For
clear-skies, the trend in winter is \(0.4\,\%/y\), larger than for
all-skies (\(0.7\,\%/y\)), which is another indication of a decreasing
trend in cloud optical thickness. Moreover, the trends under clear- and
cloudy-sky conditions are almost complementary to each other,
particularly for winter and autumn, where the signal is stronger. During
spring and summer the statistical significance is very low and the
actual trend too small for a meaningful comparison.

@@

\caption{\label{tab:trendseasontable}SDR linear trends of monthly anomalies for each season of the year.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\centering}X>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\centering}X}
\toprule
Trend [\%/year] & slope.p & cor.estimate & Sky condition & Season & t\_eff & t\_eff\_cri & conf\_2.5 & conf\_97.5\\
\midrule
\cellcolor{gray!6}{0.7000} & \cellcolor{gray!6}{0.002770} & \cellcolor{gray!6}{0.5353286} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{2e-07} & \cellcolor{gray!6}{2.060774} & \cellcolor{gray!6}{-0.0011} & \cellcolor{gray!6}{0.0015253}\\
\cmidrule{1-9}
0.1070 & 0.371000 & 0.1663281 & All skies & Spring & 1e-07 & 2.011940 & -0.0011 & 0.0015253\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.1050} & \cellcolor{gray!6}{0.175000} & \cellcolor{gray!6}{0.2543074} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{0e+00} & \cellcolor{gray!6}{2.050358} & \cellcolor{gray!6}{-0.0011} & \cellcolor{gray!6}{0.0015253}\\
\cmidrule{1-9}
0.4230 & 0.009390 & 0.4663313 & All skies & Autumn & 1e-07 & 2.074143 & -0.0011 & 0.0015253\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.3950} & \cellcolor{gray!6}{0.000515} & \cellcolor{gray!6}{0.5956771} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{0e+00} & \cellcolor{gray!6}{2.116842} & \cellcolor{gray!6}{-0.0011} & \cellcolor{gray!6}{0.0015253}\\
\cmidrule{1-9}
0.0591 & 0.497000 & 0.1267072 & Clear skies & Spring & 0e+00 & 2.069495 & -0.0011 & 0.0015253\\
\cmidrule{1-9}
\cellcolor{gray!6}{-0.0540} & \cellcolor{gray!6}{0.106000} & \cellcolor{gray!6}{-0.3013716} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{0e+00} & \cellcolor{gray!6}{2.035625} & \cellcolor{gray!6}{-0.0011} & \cellcolor{gray!6}{0.0015253}\\
\cmidrule{1-9}
0.0539 & 0.366000 & 0.1712676 & Clear skies & Autumn & 0e+00 & 2.119437 & -0.0011 & 0.0015253\\
\cmidrule{1-9}
\cellcolor{gray!6}{0.7590} & \cellcolor{gray!6}{0.000665} & \cellcolor{gray!6}{0.5861420} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Winter} & \cellcolor{gray!6}{1e-07} & \cellcolor{gray!6}{2.077321} & \cellcolor{gray!6}{-0.0011} & \cellcolor{gray!6}{0.0015253}\\
\cmidrule{1-9}
0.0621 & 0.593000 & 0.0997924 & Cloudy skies & Spring & 1e-07 & 2.026109 & -0.0011 & 0.0015253\\
\cmidrule{1-9}
\cellcolor{gray!6}{-0.0789} & \cellcolor{gray!6}{0.560000} & \cellcolor{gray!6}{-0.1109085} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Summer} & \cellcolor{gray!6}{1e-07} & \cellcolor{gray!6}{2.013800} & \cellcolor{gray!6}{-0.0011} & \cellcolor{gray!6}{0.0015253}\\
\cmidrule{1-9}
0.1910 & 0.384000 & 0.1648275 & Cloudy skies & Autumn & 1e-07 & 2.055579 & -0.0011 & 0.0015253\\
\bottomrule
\end{tabu}
\end{table}

@@

We have demonstrated that in the period 1993 - 2023, there is a positive
trend in SDR of (\(0.38\,\%/y\)) (brightening) (positive trend) in
Thessaloniki, Greece, under all-sky conditions. A previous study
\citep{Bais2013} for the period 1993 - 2011 found also a positive trend
of \(0.33\,\%/y\). The increase of this trend indicates that the
brightening of SDR continues and is probably caused by continuing
decreases in aerosol optical depth and the optical thickness of clouds
over the area. Moreover, we found a similar trend under clear-sky
conditions (\(0.097\,\%/y\)) that further supports the assumption that
the brightening is caused mainly by decreasing aerosols. Unfortunately,
for the entire period there is no available data for the aerosols, in
order to quantify their effect on SDR. However, \citet{Siomos2020} have
shown that aerosol optical depth over Thessaloniki is decreasing
constantly at least up to 2018. The attenuation of SDR by aerosols over
Europe have been proposed as major factor by \citet{Wild2021}. The
dimming effect on SDR under cloudy-sky conditions (\(0.41\,\%/y\)),
suggests that cloud optical thickness is decreasing during this period.
Because we have no adequate data to investigate the long term changes of
cloud thickness in the region, we cannot verify if the negative SDR
trend we observe under under cloudy-skies can be attributed solely to
changes in clouds.

>>>>

We calculated the linear trends of SDR, from the departures of the mean
daily values from the daily climatology and for the three sky conditions
(Table~\ref{tab:trendtable}). In Figure~\ref{fig:trendALL} we present
only the time series under all-sky conditions; the plots for clear-sky
and cloud-sky conditions, are shown in the Appendix
(Figures~\ref{fig:trendCLEAR} and~ \ref{fig:trendCLOUD}). In the studied
period, there is no significant break or change in the variability
pattern of the time series. The linear trends in all three datasets are
positive and around \(0.4\,\%/y\) for all-sky and cloudy-sky conditions,
while for clear-skies the trend is much smaller
(\textasciitilde{}\(0.41\,\%/y\)). The linear trends were calculated
taking into account the autocorrelation of the time series and all three
are statistically significant at least at the \(95\,\%\) confidence
level, as they are larger than the corresponding \(2\sigma\) errors. The
clear-sky trend is very small suggesting a small effect from aerosols
and water vapor which are the dominant factors of the SDR variability.
In contrast the large positive trend of SDR under cloudy skies can be
attributed to reduction in cloud cover and/or optical depth. Lack of
continuous cloud observations that could support these findings does not
allow drawing firm conclusions. However, there are indications that the
total cloud-cover as inferred from the ERA5 analysis for the grid point
of Thessaloniki is decreasing over the period of study.

The all-sky trend is similar to the one reported in \citet{Bais2013}
from a ten-year shorter dataset suggesting that the tendency of SDR in
Thessaloniki is systematic. Other studies for the European region
reported a change in the SDR trend around 1980 from negative to positive
with comparable magnitude \citep{Wild2021, Yuan2021, Ohmura2009}, well
before the start of our records. However, the trends reported here for
the three datasets are in accordance with the widely accepted solar
radiation brightening over Europe. For the period of our observations
the trend in the TSI is negligible (\(-0.00024\,\%/y\)), and thus we
cannot attribute any significant effect on the SDR trend to solar
variability.

@@

For cloudy-sky conditions (Figure~\ref{fig:szatrends-3}), we can not
discern any significant dependence of the SDR trend with SZA as the
variability of irradiance is dominated by the cloud effects. For SZAs
\(25^\circ\) -- \(75^\circ\), the trends are variable and for most SZAs
insignificant ranging between \(0\) and \(0.2\,\%/y\). The smooth
increase of the trend at SZAs larger than \(\sim 75^\circ\) is probably
associated with stronger attenuation by clouds under oblique incidence
angles which results in smaller variability.

@@

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend for the three sky conditions, and
for each season of the year, using the corresponding mean monthly
anomalies (Figure~\ref{fig:seasonalALL} and
Table~\ref{tab:trendseasontable}). For all-sky conditions the trend in
SDR in winter is the largest (\(0.7\,\%/y\)), followed by the trend in
autumn (\(0.42\,\%/y\), a value close to the long term trend) both
statistically significant above the \(99\,\%\) confidence level. In
spring and summer, the trends are much smaller and of lesser statistical
significance. These seasonal differences indicate a possible relation of
the trends in SDR to trends of clouds during winter and autumn. For
clear-skies, the trend in winter is \(0.4\,\%/y\), larger than for
all-skies (\(0.7\,\%/y\)), which is another indication of a decreasing
trend in cloud optical thickness. Moreover, the trends under clear- and
cloudy-sky conditions are almost complementary to each other,
particularly for winter and autumn, where the signal is stronger. During
spring and summer the statistical significance is very low and the
actual trend too small for a meaningful comparison.

@@

\caption{\label{tab:trendseasontable}SDR linear trends of monthly anomalies for each season of the year.}
\begin{tabu} to \linewidth {>{\centering\arraybackslash}p{8em}>{\centering}X>{\raggedleft}X>{\raggedleft}X>{\centering}X}
\toprule
Trend [\%/year] & slope.p & Pearson cor. & Sky condition & Season\\
\midrule
\cellcolor{gray!6}{0.7000} & \cellcolor{gray!6}{0.002770} & \cellcolor{gray!6}{0.5350} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Winter}\\
\cmidrule{1-5}
0.1070 & 0.371000 & 0.1660 & All skies & Spring\\
\cmidrule{1-5}
\cellcolor{gray!6}{0.1050} & \cellcolor{gray!6}{0.175000} & \cellcolor{gray!6}{0.2540} & \cellcolor{gray!6}{All skies} & \cellcolor{gray!6}{Summer}\\
\cmidrule{1-5}
0.4230 & 0.009390 & 0.4660 & All skies & Autumn\\
\cmidrule{1-5}
\cellcolor{gray!6}{0.3950} & \cellcolor{gray!6}{0.000515} & \cellcolor{gray!6}{0.5960} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Winter}\\
\cmidrule{1-5}
0.0591 & 0.497000 & 0.1270 & Clear skies & Spring\\
\cmidrule{1-5}
\cellcolor{gray!6}{-0.0540} & \cellcolor{gray!6}{0.106000} & \cellcolor{gray!6}{-0.3010} & \cellcolor{gray!6}{Clear skies} & \cellcolor{gray!6}{Summer}\\
\cmidrule{1-5}
0.0539 & 0.366000 & 0.1710 & Clear skies & Autumn\\
\cmidrule{1-5}
\cellcolor{gray!6}{0.7590} & \cellcolor{gray!6}{0.000665} & \cellcolor{gray!6}{0.5860} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Winter}\\
\cmidrule{1-5}
0.0621 & 0.593000 & 0.0998 & Cloudy skies & Spring\\
\cmidrule{1-5}
\cellcolor{gray!6}{-0.0789} & \cellcolor{gray!6}{0.560000} & \cellcolor{gray!6}{-0.1110} & \cellcolor{gray!6}{Cloudy skies} & \cellcolor{gray!6}{Summer}\\
\cmidrule{1-5}
0.1910 & 0.384000 & 0.1650 & Cloudy skies & Autumn\\
\bottomrule
\end{tabu}
\end{table}

@@

We have demonstrated that in the period 1993 - 2023, there is a positive
trend in SDR of (\(0.38\,\%/y\)) (brightening) (positive trend) in
Thessaloniki, Greece, under all-sky conditions. A previous study
\citep{Bais2013} for the period 1993 - 2011 found also a positive trend
of \(0.33\,\%/y\). The increase of this trend indicates that the
brightening of SDR continues and is probably caused by continuing
decreases in aerosol optical depth and the optical thickness of clouds
over the area. Moreover, we found a similar trend under clear-sky
conditions (\(0.1\,\%/y\)) that further supports the assumption that the
brightening is caused mainly by decreasing aerosols. Unfortunately, for
the entire period there is no available data for the aerosols, in order
to quantify their effect on SDR. However, \citet{Siomos2020} have shown
that aerosol optical depth over Thessaloniki is decreasing constantly at
least up to 2018. The attenuation of SDR by aerosols over Europe have
been proposed as major factor by \citet{Wild2021}. The dimming effect on
SDR under cloudy-sky conditions (\(0.41\,\%/y\)), suggests that cloud
optical thickness is decreasing during this period. Because we have no
adequate data to investigate the long term changes of cloud thickness in
the region, we cannot verify if the negative SDR trend we observe under
under cloudy-skies can be attributed solely to changes in clouds.
